Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-01-20 paper/drinking-bioinformatics.tex

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.
\begin{figure*}
  \centering
%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \includegraphics[width=1\textwidth]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

\bibliographystyle{natbib}
\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

>>>>

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.eps}
  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.
\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth]{overview.eps}
%  \includegraphics[width=1\textwidth]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

\bibliographystyle{natbib}
\begin{thebibliography}{}

\bibitem[Ashburner {\em et~al.}(2000)Ashburner, Ball, Blake, Botstein, Butler,
  Cherry, Davis, Dolinski, Dwight, Eppig, Harris, Hill, Issel-Tarver,
  Kasarskis, Lewis, Matese, Richardson, Ringwald, Rubin, and
  Sherlock]{Ashburner2000b}
Ashburner, M., Ball, C.~A., Blake, J.~A., Botstein, D., Butler, H., Cherry,
  J.~M., Davis, A.~P., Dolinski, K., Dwight, S.~S., Eppig, J.~T., Harris,
  M.~A., Hill, D.~P., Issel-Tarver, L., Kasarskis, A., Lewis, S., Matese,
  J.~C., Richardson, J.~E., Ringwald, M., Rubin, G.~M., and Sherlock, G.
  (2000).
\newblock Gene ontology: tool for the unification of biology. the gene ontology
  consortium.
\newblock {\em Nat Genet\/}, {\bf 25}(1), 25--29.

\bibitem[Bada and Hunter(2007)Bada and Hunter]{Bada2007a}
Bada, M. and Hunter, L. (2007).
\newblock Enrichment of obo ontologies.
\newblock {\em Journal of Biomedical Informatics\/}, {\bf 40}(3), 300--315.

\bibitem[Bada {\em et~al.}(2004)Bada, Stevens, Goble, Gil, Ashburner, Blake,
  Cherry, Harris, and Lewis]{Bada2004}
Bada, M., Stevens, R., Goble, C., Gil, Y., Ashburner, M., Blake, J.~A., Cherry,
  M.~J., Harris, M., and Lewis, S. (2004).
\newblock A short study on the success of the gene ontology.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web\/}, {\bf 1}(2), 235--240.

\bibitem[Bard {\em et~al.}(2005)Bard, Rhee, and Ashburner]{Bard2005}
Bard, J., Rhee, S.~Y., and Ashburner, M. (2005).
\newblock An ontology for cell types.
\newblock {\em Genome Biology\/}, {\bf 6}(2).

\bibitem[Berners-Lee {\em et~al.}(2001)Berners-Lee, Hendler, Lassila, {\em
  et~al.}]{Berners-Lee2001}
Berners-Lee, T., Hendler, J., Lassila, O., {\em et~al.} (2001).
\newblock {The Semantic Web}.
\newblock {\em Scientific American\/}, {\bf 284}(5), 28--37.

\bibitem[Conrad {\em et~al.}(2011)Conrad, W\"{u}nsche, Tan, Bulkescher,
  Sieckmann, Verissimo, Edelstein, Walter, Liebel, Pepperkok, and
  Ellenberg]{Conrad2011}
Conrad, C., W\"{u}nsche, A., Tan, T. H.~H., Bulkescher, J., Sieckmann, F.,
  Verissimo, F., Edelstein, A., Walter, T., Liebel, U., Pepperkok, R., and
  Ellenberg, J. (2011).
\newblock Micropilot: automation of fluorescence microscopy-based imaging for
  systems biology.
\newblock {\em Nature methods\/}, {\bf 8}(3), 246--249.

\bibitem[Cook {\em et~al.}(2011)Cook, Bookstein, and Gennari]{opb}
Cook, D.~L., Bookstein, F.~L., and Gennari, J.~H. (2011).
\newblock Physical properties of biological entities: An introduction to the
  ontology of physics for biology.
\newblock {\em PLoS ONE\/}, {\bf 6}(12), e28708.

\bibitem[Eilbeck {\em et~al.}(2005)Eilbeck, Lewis, Mungall, Yandell, Stein,
  Durbin, and Ashburner]{Eilbeck2005}
Eilbeck, K., Lewis, S.~E., Mungall, C.~J., Yandell, M., Stein, L., Durbin, R.,
  and Ashburner, M. (2005).
\newblock The sequence ontology: A tool for the unification of genome
  annotations.
\newblock {\em Genome Biology\/}, {\bf 6}(R55).

\bibitem[Engel {\em et~al.}(2010)Engel, Balakrishnan, Binkley, Christie,
  Costanzo, Dwight, Fisk, Hirschman, Hitz, Hong, Krieger, Livstone, Miyasato,
  Nash, Oughtred, Park, Skrzypek, Weng, Wong, Dolinski, Botstein, and
  Cherry]{ypo}
Engel, S.~R., Balakrishnan, R., Binkley, G., Christie, K.~R., Costanzo, M.~C.,
  Dwight, S.~S., Fisk, D.~G., Hirschman, J.~E., Hitz, B.~C., Hong, E.~L.,
  Krieger, C.~J., Livstone, M.~S., Miyasato, S.~R., Nash, R., Oughtred, R.,
  Park, J., Skrzypek, M.~S., Weng, S., Wong, E.~D., Dolinski, K., Botstein, D.,
  and Cherry, J.~M. (2010).
\newblock {Saccharomyces Genome Database provides mutant phenotype data}.
\newblock {\em Nucleic Acids Research\/}, {\bf 38}(suppl 1), D433--D436.

\bibitem[Fuchs {\em et~al.}(2010)Fuchs, Pau, Kranz, Sklyar, Budjan, Steinbrink,
  Horn, Pedal, Huber, and Boutros]{Fuchs2010}
Fuchs, F., Pau, G., Kranz, D., Sklyar, O., Budjan, C., Steinbrink, S., Horn,
  T., Pedal, A., Huber, W., and Boutros, M. (2010).
\newblock Clustering phenotype populations by genome-wide {RNAi} and
  multiparametric imaging.
\newblock {\em Molecular Systems Biology\/}, {\bf 6}.

\bibitem[{Gene Ontology Consortium}(2010){Gene Ontology Consortium}]{go2010}
{Gene Ontology Consortium} (2010).
\newblock The gene ontology in 2010: extensions and refinements.
\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue), D331--335.

\bibitem[Gkoutos and Hoehndorf(2011)Gkoutos and Hoehndorf]{obml2011h1}
Gkoutos, G.~V. and Hoehndorf, R. (2011).
\newblock Ontology-based cross-species integration and analysis of
  saccharomyces cerevisiae phenotypes.
\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
  and Life sciences (OBML)\/}.

\bibitem[Gkoutos {\em et~al.}(2005)Gkoutos, Green, Mallon, Hancock, and
  Davidson]{Gkoutos2005}
Gkoutos, G.~V., Green, E.~C., Mallon, A.-M.~M., Hancock, J.~M., and Davidson,
  D. (2005).
\newblock {Using ontologies to describe mouse phenotypes.}
\newblock {\em Genome biology\/}, {\bf 6}(1).

\bibitem[Gkoutos {\em et~al.}(2009)Gkoutos, Mungall, Dolken, Ashburner, Lewis,
  Hancock, Schofield, Kohler, and Robinson]{Gkoutos2009b}
Gkoutos, G.~V., Mungall, C., Dolken, S., Ashburner, M., Lewis, S., Hancock, J.,
  Schofield, P., Kohler, S., and Robinson, P.~N. (2009).
\newblock Entity/quality-based logical definitions for the human skeletal
  phenome using {PATO}.
\newblock {\em Annual International Conference of the IEEE Engineering in
  Medicine and Biology Society.}, {\bf 1}, 7069--7072.

\bibitem[Goble and Stevens(2008)Goble and Stevens]{goble}
Goble, C. and Stevens, R. (2008).
\newblock State of the nation in data integration for bioinformatics.
\newblock {\em Journal of Biomedical Informatics\/}, {\bf 41}(5), 687--693.

\bibitem[Grau {\em et~al.}(2008)Grau, Horrocks, Motik, Parsia, Patelschneider,
  and Sattler]{Grau2008}
Grau, B., Horrocks, I., Motik, B., Parsia, B., Patelschneider, P., and Sattler,
  U. (2008).
\newblock {OWL} 2: The next step for {OWL}.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web\/}, {\bf 6}(4), 309--322.

\bibitem[Gruber(1995)Gruber]{Gruber1995}
Gruber, T.~R. (1995).
\newblock Toward principles for the design of ontologies used for knowledge
  sharing.
\newblock {\em International Journal of Human-Computer Studies\/}, {\bf
  43}(5-6).

\bibitem[Guarino(1998)Guarino]{Guarino1998}
Guarino, N. (1998).
\newblock Formal ontology and information systems.
\newblock In N.~Guarino, editor, {\em Proceedings of the 1st International
  Conference on Formal Ontologies in Information Systems\/}, pages 3--15. IOS
  Press.

\bibitem[Herre {\em et~al.}(2006)Herre, Heller, Burek, Hoehndorf, Loebe, and
  Michalek]{Herre2006}
Herre, H., Heller, B., Burek, P., Hoehndorf, R., Loebe, F., and Michalek, H.
  (2006).
\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
  integrating objects and processes [{V}ersion 1.0].
\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany.

\bibitem[Hoehndorf {\em et~al.}(2010a)Hoehndorf, Oellrich, and
  Rebholz-Schuhmann]{Hoehndorf2010phene}
Hoehndorf, R., Oellrich, A., and Rebholz-Schuhmann, D. (2010a).
\newblock Interoperability between phenotype and anatomy ontologies.
\newblock {\em Bioinformatics\/}, {\bf 26}(24), 3112 -- 3118.

\bibitem[Hoehndorf {\em et~al.}(2010b)Hoehndorf, Oellrich, Dumontier, Kelso,
  Rebholz-Schuhmann, and Herre]{Hoehndorf2010patterns}
Hoehndorf, R., Oellrich, A., Dumontier, M., Kelso, J., Rebholz-Schuhmann, D.,
  and Herre, H. (2010b).
\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
\newblock {\em BMC Bioinformatics\/}, {\bf 11}(1), 441+.

\bibitem[Horridge {\em et~al.}(2007)Horridge, Bechhofer, and
  Noppens]{Horridge2007}
Horridge, M., Bechhofer, S., and Noppens, O. (2007).
\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
  OWL Experiences and Directions\/}.

\bibitem[Horrocks(2007)Horrocks]{Horrocks2007}
Horrocks, I. (2007).
\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
  {W}eb {O}ntology {L}anguage.
\newblock Technical report, University of Manchester.
\newblock \url{http://www.cs.man.ac.uk/~horrocks/obo/}.

\bibitem[Kazakov {\em et~al.}(2011)Kazakov, Kr{\"o}tzsch, and
  Siman\v{c}\'{i}k]{Kazakov2011}
Kazakov, Y., Kr{\"o}tzsch, M., and Siman\v{c}\'{i}k, F. (2011).
\newblock Unchain my $\mathcal{EL}$ reasoner.
\newblock In {\em Proceedings of the 23rd International Workshop on Description
  Logics (DL'10)\/}, CEUR Workshop Proceedings. CEUR-WS.org.

\bibitem[Lock and Strömblad(2010)Lock and Strömblad]{Lock2010}
Lock, J.~G. and Strömblad, S. (2010).
\newblock Systems microscopy: an emerging strategy for the life sciences.
\newblock {\em Experimental Cell Research\/}, {\bf 316}(8), 1438--1444.

\bibitem[Mungall {\em et~al.}(2010a)Mungall, Gkoutos, Smith, Haendel, Lewis,
  and Ashburner]{Mungall2010}
Mungall, C., Gkoutos, G., Smith, C., Haendel, M., Lewis, S., and Ashburner, M.
  (2010a).
\newblock Integrating phenotype ontologies across multiple species.
\newblock {\em Genome Biology\/}, {\bf 11}(1), R2+.

\bibitem[Mungall {\em et~al.}(2010b)Mungall, Bada, Berardini, Deegan, Ireland,
  Harris, Hill, and Lomax]{Mungall2010go}
Mungall, C.~J., Bada, M., Berardini, T.~Z., Deegan, J., Ireland, A., Harris,
  M.~A., Hill, D.~P., and Lomax, J. (2010b).
\newblock Cross-product extensions of the gene ontology.
\newblock {\em Journal of biomedical informatics\/}.
\newblock in press.

\bibitem[Neumann {\em et~al.}(2010)Neumann, Walter, Hériché, Bulkescher,
  Erfle, Conrad, Rogers, Poser, Held, Liebel, and et~al.]{Neumann2010}
Neumann, B., Walter, T., Hériché, J.-K., Bulkescher, J., Erfle, H., Conrad,
  C., Rogers, P., Poser, I., Held, M., Liebel, U., and et~al. (2010).
\newblock Phenotypic profiling of the human genome by time-lapse microscopy
  reveals cell division genes.
\newblock {\em Nature\/}, {\bf 464}(7289), 721--727.

\bibitem[Ogren {\em et~al.}(2004)Ogren, Cohen, Acquaah-Mensah, Eberlein, and
  Hunter]{Ogren2004}
Ogren, P.~V., Cohen, K.~B., Acquaah-Mensah, G.~K., Eberlein, J., and Hunter, L.
  (2004).
\newblock The compositional structure of gene ontology terms.
\newblock {\em Pac Symp Biocomput\/}, pages 214--225.

\bibitem[\"{O}zg\"{o}vde and Gr\"{u}ninger(2010)\"{O}zg\"{o}vde and
  Gr\"{u}ninger]{Gruninger2010}
\"{O}zg\"{o}vde, A. and Gr\"{u}ninger, M. (2010).
\newblock Foundational process relations in bio-ontologies.
\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
  Information Systems\/}, pages 243--256, Amsterdam, The Netherlands, The
  Netherlands. IOS Press.

\bibitem[Robinson {\em et~al.}(2008)Robinson, Koehler, Bauer, Seelow, Horn, and
  Mundlos]{Robinson2008}
Robinson, P.~N., Koehler, S., Bauer, S., Seelow, D., Horn, D., and Mundlos, S.
  (2008).
\newblock The human phenotype ontology: a tool for annotating and analyzing
  human hereditary disease.
\newblock {\em American journal of human genetics\/}, {\bf 83}(5), 610--615.

\bibitem[Schindelman {\em et~al.}(2011)Schindelman, Fernandes, Bastiani, Yook,
  and Sternberg]{wpo}
Schindelman, G., Fernandes, J., Bastiani, C., Yook, K., and Sternberg, P.
  (2011).
\newblock Worm phenotype ontology: integrating phenotype data within and beyond
  the c. elegans community.
\newblock {\em BMC Bioinformatics\/}, {\bf 12}(1), 32.

\bibitem[Schmitz {\em et~al.}(2010)Schmitz, Held, Janssens, Hutchins, Hudecz,
  Ivanova, Goris, Trinkle-Mulcahy, Lamond, Poser, Hyman, Mechtler, Peters, and
  Gerlich]{Schmitz2010}
Schmitz, M. H.~A., Held, M., Janssens, V., Hutchins, J. R.~A., Hudecz, O.,
  Ivanova, E., Goris, J., Trinkle-Mulcahy, L., Lamond, A.~I., Poser, I., Hyman,
  A.~A., Mechtler, K., Peters, J.-M., and Gerlich, D.~W. (2010).
\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
  importin-$\beta$1 as key mitotic exit regulators in human cells.
\newblock {\em Nature Cell Biology\/}, {\bf 12}, 886--893.

\bibitem[Schofield {\em et~al.}(2011)Schofield, Sundberg, Hoehndorf, and
  Gkoutos]{Schofield2011}
Schofield, P.~N., Sundberg, J.~P., Hoehndorf, R., and Gkoutos, G.~V. (2011).
\newblock New approaches to the representation and analysis of phenotype
  knowledge in human diseases and their animal models.
\newblock {\em Briefings in Functional Genomics\/}, {\bf 10}(5), 258--265.

\bibitem[Smith {\em et~al.}(2007)Smith, Ashburner, Rosse, Bard, Bug, Ceusters,
  Goldberg, Eilbeck, Ireland, Mungall, Leontis, Serra, Ruttenberg, Sansone,
  Scheuermann, Shah, Whetzel, and Lewis]{Smith2007}
Smith, B., Ashburner, M., Rosse, C., Bard, J., Bug, W., Ceusters, W., Goldberg,
  L.~J., Eilbeck, K., Ireland, A., Mungall, C.~J., Leontis, N., Serra, P.~R.,
  Ruttenberg, A., Sansone, S.~A., Scheuermann, R.~H., Shah, N., Whetzel, P.~L.,
  and Lewis, S. (2007).
\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
  biomedical data integration.
\newblock {\em Nat Biotech\/}, {\bf 25}(11), 1251--1255.

\bibitem[Smith {\em et~al.}(2004)Smith, Goldsmith, and Eppig]{Smith2004}
Smith, C.~L., Goldsmith, C.-A.~W., and Eppig, J.~T. (2004).
\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
  and comparing phenotypic information.
\newblock {\em Genome Biology\/}, {\bf 6}(1), R7.

\end{thebibliography}

%\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
